---
title: "Herminium monorchis root and seed fungi"
author: "Norwegian institute for nature research"
date: "10/03/2024"
format: pdf
pdf-engine: pdflatex
---

This script treats data collected from root and seed capsules of Herminium morchis at two sites on Hvaler, Norway.

Before getting started, we can run this line to make sure R can read Norwegian letters.

```{r}
#| label: Language settings
#| warning: false

Sys.setlocale(locale='no_NB.utf8')

```

Load the required packages.

```{r}
#| label: Install packages
#| warning: false

library(tidyverse)
library(stringr)
library(readxl)


#NB use install.packages("package name") if the package is not installed yet.
#Using qmd requires the "rmarkdown" package installed.
```

Then we can import the data

```{r}
#| label: Import data
#| warning: false

df <- read_excel("P:/154027_effektovervaking_av_trua_arter_og_naturtyper_2024/04 Honningblom/Herminium monorchis/Data/Artslist_rot_og_fro_04042024.xlsx", sheet = 'Sheet1')
```

The data frame contains some columns we do not need. We can also give them a better name.

```{r}
#| label: Select relevant columns
#| warning: false

df <- df %>% 
  select(c("final_phylum", 
           "final_class", 
           "final_order", 
           "final_family", 
           "final_genus", 
           "final_species", 
           "komment_fra_Marie", 
           "root1":"SKJ_F5"))

df <- df %>% 
  rename("Phylum" = "final_phylum", 
           "Class" = "final_class", 
           "Order" = "final_order", 
           "Family" = "final_family", 
           "Genus" = "final_genus", 
           "Species" = "final_species", 
           "Comment" = "komment_fra_Marie", 
           "SKJ_R1" = "root1", 
           "SKJ_R2" = "root2")
```

Now we can convert the dataset from a wide to long format.

```{r}
#| label: Create long data set
#| warning: false

df_long <- df %>% pivot_longer(cols = "SKJ_R1":"SKJ_F5", names_to = "Sample_ID", values_to = "Count")
```

Add a column that describes the locality and sample type.

```{r}
#| label: Include locality, sample type
#| warning: false

df_long <- df_long %>% mutate(
    Locality = case_when(str_detect(Sample_ID, "SKJ") ~ "Skjellvik", 
                       str_detect(Sample_ID, "SKI") ~ "Skipstadsand"), 
    Sample_type = case_when(str_detect(Sample_ID, "R") ~ "root",
                   str_detect(Sample_ID, "F") ~ "seed_capsule"))
```

Export the dataset as xlsx.

```{r}
#| label: Export data
#| warning: false

#Select the columns we need, in the proper order. 
df_long <- df_long %>% select(Locality,
                              Sample_ID, 
                              Sample_type, 
                              Phylum,
                              Class, 
                              Order,
                              Family, 
                              Genus, 
                              Species, 
                              Count, 
                              Comment)

#Export the dataset, best to export as .xlsx as this is most likely to preserve norwegian letters
openxlsx::write.xlsx(df_long, file = "P:/154027_effektovervaking_av_trua_arter_og_naturtyper_2024/04 Honningblom/Herminium monorchis/Data/Root_and_seed_fungi_clean_03.10.2024.xlsx", colNames = TRUE)


```
