---
title: "Script_Herminium_vegetation_composition"
author: "Norwegian institute for nature research"
format: html
editor: visual
---

## About

This script is to analyse vegetation composition in plots with and without Herminium monorchis.

## Language settings

Before getting started, we can run this line to make sure R can read Norwegian letters.

```{r}
#| label: Language settings   
#| warning: false  
Sys.setlocale(locale='no_NB.utf8')
```

## Packages

Load required packages.

```{r}
 #| label: Language settings 
 #| warning: false

library(reshape2)
library(readxl)
library(tidyverse)
library(vegan)
library(ggrepel)
```

## Load data

```{r}
#| label: Load data
#| warning: false

df <- read_xlsx("P:/154027_effektovervaking_av_trua_arter_og_naturtyper_2024/04 Honningblom/Herminium monorchis/Data/Herminium_monorchis_masterfile_v_2024.xlsx", sheet = "Plot_plants")

```

## Explore and clean data

```{r}
#| label: Exploring data
#| warning: false

str(df)
unique(df$Locality)
unique(df$Species_norwegian)

df %>%  
    pull(Species_norwegian) %>% 
    unique %>% 
    sort

#Vill-lin and Vill-lin are duplicates. Change Rapp sp. to Rappslekta, Vikke sp. til Vikkeslekta

df <- df %>% mutate(Species_norwegian = recode(Species_norwegian, 
                                         `Rapp sp.` = "Rappslekta",
                                         `Vikke sp.` = "Vikkeslekta", 
                                         `Vill-Lin` = "Vill-lin"))

#Remove rows where nothing was registered in the Quadrats. 
df <- df %>%
  filter(!is.na(Presence_quadrat))

#Remove unknown species
df <- df %>% 
  filter(!grepl('Ukjent', Species_norwegian))

#Filletassen plot 4 has a musk orchid observation that is new this year. Therefore, the plot was not sampled for vegetation composition. Removing this plot is the best option. 

df <- df %>% 
  filter(!grepl('FI4', Plot_ID_new))

#Adding a column for presence of Musk orchid

df <- df %>% 
  group_by(Plot_ID_new) %>% 
  mutate(Presence_H_monorchis = ifelse(any(Species_norwegian == "Honningblom" & Quadrats >= 1), "Present", "Absent"))

```

## Perform NMDS analysis

First, we need to prepare the data for NMDS, then perform the NMDS and make a plot. Following this and previous scripts: <https://rpubs.com/CPEL/NMDS>

```{r}
#| Label: Prepare data for NMDS
#| warning: false

#We need to create a species matrix
df_nmds <- dcast(df, Plot_ID_new + Locality + Presence_H_monorchis ~ Species_norwegian, value.var = "Quadrats",
            fun.aggregate = sum, fill = 0)

#Create species dataset
df_species <- df_nmds %>% 
  select(!c(Plot_ID_new, Locality, Presence_H_monorchis))

#Create grouping dataset
df_groups <- df_nmds %>% 
  select(Plot_ID_new, Locality, Presence_H_monorchis)

# Calculating relative abundance and creating new dataframe with relative abundance data
df_species_rel <-         
  decostand(df_species, method = "total")

# Calculating distances
df_distmat <- 
  vegdist(df_species_rel, method = "bray")

# Create matrix
df_distmat <- 
  as.matrix(df_distmat, labels = T)
#write.csv(island.spp_distmat, "island_spp_distmat.csv")

# Running NMDS
NMDS <-
  metaMDS(df_distmat,
          distance = "bray",
          k = 3,
          maxit = 999, 
          trymax = 500,
          wascores = TRUE)

stressplot(NMDS)#Creates stress plot

#Making a quick plot
plot(NMDS, "sites")

#Extract NMDS scores and combine with grouping variables
df_nmds_scores = as.data.frame(scores(NMDS))
df_nmds_scores <- bind_cols(df_groups, df_nmds_scores)

#Making a nice ggplot

NMDS.plot <- ggplot(df_nmds_scores, aes(x = NMDS1, y = NMDS2))+
  geom_point(size = 4, aes(colour = Locality, shape = as.factor(Presence_H_monorchis)))+
  theme(axis.text.y = element_text(size = 12), 
        axis.text.x = element_text(size = 12), 
        legend.text = element_text(size = 12), 
        legend.position = "right") +
  theme_classic()+
  labs(x = "NMDS1", y = "NMDS2") +
  geom_text_repel(aes(label = Plot_ID_new), size=2)+
  guides(colour = guide_legend(title="Lokalitet"), 
         shape = guide_legend("Forekomst i ruten"))+
  scale_shape_manual(values = c(21,19), 
                     labels = c("Uten", "Med"))

NMDS.plot

ggsave("P:/154027_effektovervaking_av_trua_arter_og_naturtyper_2024/04 Honningblom/Herminium monorchis/Output/NMDS_honningblom.jpeg", NMDS.plot, width = 16, height = 12, units = "cm")


```
